---
title: "Workshop: Intermediate R Software Development"
author: "Martin Morgan"
date: "5 - 21 October , 2016"
vignette: >
  %\VignetteIndexEntry{Workshop: Intermediate R Software Development}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, echo=FALSE}
suppressPackageStartupMessages({
    library(devtools)
    library(roxygen2)
})
```

This package grew out of a series of meetings to enhance our
understanding of _R_ as a programming language, and to connect with
other people interested in _R_ software development. It was open to
members of the Roswell Park and SUNY at Buffalo communities. It is
suitable for people who already know _R_, and are eager to develop a
deeper understanding of the language and 'best practices' when
tackling larger projects.

# 1. Working with packages and github

The main objectives of this session are to write an [R][] package, and
to use [github][] for version control.

## _R_ packages

An _R_ package is nothing more than an on-disk folder contains
specific files and folders. A minimal package contains

1. A text `DESCRIPTION` file with metadata (e.g., title, author,
   description, license, package dependencies).
2. A text `NAMESPACE` file. The NAMESPACE file describes the packages,
   functions, methods, and classes imported (used by) the _R_ code in
   the package. The NAMESPACE file also specifies which functions,
   classes, methods, and data defined in the package are visible to
   the user. A minimal NAMESPACE file contains nothing -- no symbols
   are used from other packages, and no symbols defined in the package
   are visible to the user.
   
### Exercise

Create a nearly minimal _R_ package using the [devtools][] function
`create`. We'll create the pacakge in a temporary directory, but do
this in a permanent location.

```{r}
pkg <- file.path(tempdir(), "Elbo")
devtools::create(pkg, description=list(Version="0.0.1"))
```

Check out the content of the directory

```{r}
dir(pkg, recursive=TRUE)
```

The `Elbo.Rproj` file is for use with [RStudio][], and is
optional. The content of the `DESCRIPTION` file is

```{r}
cat(strwrap({
    readLines(file.path(pkg, "DESCRIPTION"))
}, exdent=4), sep="\n")
```

Usually, one would edit the file:

- The Title: and Description: fields to describe the package.
- The Authors@R: field provides a facility for enumerating one or
  several authors.
- The Version: field is meant to be incremented with each change to
  the package; it allows users and developers to know precisely which
  version of the package is in use. The [Bioconductor][] convention is
  to use versions with format x.y.z, e.g., 0.0.1, 0.99.0, 1.0.0, 1.0.1
- The License: field is used to describe the conditions under which
  the pacakge is made available; one often chooses a standard license,
  e.g., 'Artistic-2.0' or 'GPL 3'

The package can actually be installed and loaded, e.g., using [devtools][]

```{r}
devtools::install(pkg)
library(Elbo)
```

The package is now on the search path (at the second position)

```{r}
head(search())
```

but of course has no symbols available

```{r}
ls(pos=2)
```

## Version control

During software development, one wants to be able to make changes to a
package in a way that allows one to easily record what one has done,
to revert back to a previous 'working' version if one ends up going
down the wrong path, and to share the package with colleagues. 

Version control enables each of these objects. [git][] is one version
control system, it's made especially useful through the [github][]
website.

- Make sure that [git][] is installed on your system.
- Create a user account on [github][]

Each package we create will be managed as a github
_repository_.

- [Create a github repository][1], naming the repository after your
  package.
- [Add your Elbo package][2] to the repository

There are two important steps taken when addign your package to the repository.

1. _Commit_ to the local repository, `git commit ...`. One often makes
   a number of commits to the local repository, perhaps after every
   meaningful bit of code is produced -- several times an hour, for
   instance.
2. _Push_ a series of commits to the github repository `git push
   ...`. This is often associate with the completion of a conceptual
   feature, or perhaps at the end of the day.

[1]: https://help.github.com/articles/create-a-repo/
[2]: https://help.github.com/articles/adding-an-existing-project-to-github-using-the-command-line/

## Package development: a first function

Here's a simple function

_R_ packages typically consist of _R_ functions. To add a function to
your package,

- Create a directory `R` inside the `Elbo` directory
- Create a text file `hi.R` inside the `Elbo/R` directory

```{r}
dir.create(file.path(pkg, "R"))
file.create(file.path(pkg, "R", "hi.R"))
```

We'll add the following function to the `hi.R` file:

```
hi <- function(who) {
    paste("hello", who, "you have", nchar(who), "letters in your name!")
}
```

A common practice uses the [roxygen2][] package to help document and
manage functions in packages. The idea is to include a few lines of
text above the function, using tags such as `@param` to indicate
different parts of the documentation. Update the `hi.R` package, using
a text editor of your choice, to read as follows:

```
#' Help start conversations
#'
#' This function generates some helpful text that can be used to start
#' conversations in all kinds of awkward social situations.
#'
#' @param who character(1) The name of the person you wish to start a
#'     conversation with.
#'
#' @return character(1) A line of text to be used when starting conversations.
#'
#' @examples
#' hi("Martin Morgan")
#'
#' @export
hi <- function(who) {
    paste("hello", who, "you have", nchar(who), "letters in your name!")
}
```

```{r, echo=FALSE}
txt <- c(
    "#' Help start conversations",
    "#'",
    "#' This function generates some helpful text that can be used to start",
    "#' conversations in all kinds of awkward social situations.",
    "#'",
    "#' @param who character(1) The name of the person you wish to start a",
    "#'     conversation with.",
    "#'",
    "#' @return character(1) A line of text to be used when starting conversations.",
    "#'",
    "#' @examples",
    "#' hi('Martin Morgan')",
    "#'",
    "#' @export",
    "hi <- function(who) {",
    "    paste('hello', who, 'you have', nchar(who), 'letters in your name!')",
    "}"
)
cat(txt, file=file.path(pkg, "R", "hi.R"), sep="\n")
```
Lines starting with `#'` are recognized by [roxygen2][]. The first
line in this block becomes the help page title. The next paragraph is
the `description` found on all help pages. 

- The `@param` tag is used to define arguments, e.g., the `who`
  argument is supposed to be a length 1 character vector, etc.
- `@return` describes the return value of the function.
- `@example` starts _R_ code that will be included in the `example`
  section of the man page.
- `@export` indicates that the function will be exported from the
  package, hence visible to the user.

We're now ready to compile the documentation

```{r}
devtools::document(pkg)
```

and install the updated package

```{r}
devtools::install(pkg)
```

If necessary, load the newly installed package and check out the
functionality

```{r}
library(Elbo)
head(search())
ls(pos=2)
hi("Martin")
```

Also check out our documentation

```{r}
?hi
```

## Version control!

We've made some changes, so...

- Update the version number, e.g., to 0.0.2
- Add the new files to git
- Review changes
- Commit all changed files, including the updated DESCRIPTION file
- Push the changes to github.

```
## cd Elbo
## update Version in DESCRIPTION
git add R/hi.R
git diff
git commit -a
git push
```

These operations are also available via the [RStudio][] user
interface.

### Exercise

Add an option `how` that will shout (upper case), whisper (lowercase),
or say (no change) the user name. Implement the option as separate
functions. Use `match.arg()` to select the mode of communication, and
`switch()` to return the function to be applied when creating the
return sentence. Update the documentation and add examples to the man
page.

My solution:

I implemented the code as

```
hi <- function(who, how=c("say", "shout", "whisper")) {
    how <- match.arg(how)
    fun <- switch(how, say=say, shout=shout, whisper=whisper)
    paste("hello", fun(who),  "you have", nchar(who), "letters in your name!")
}

say <- function(who) {
    who
}

shout <- function(who) {
    toupper(who)
}

whisper <- function(who) {
    tolower(who)
}
```

I updated the documentation by adding a `@param`...

```
#' @param how character(1) How to greet the conversant. Either "say"
#'     (default, no change), "shout" (upper-case), "whisper" (lower-case)
```

...and updating the `@examples`

```
#' @examples
#' hi("Martin Morgan")
#' hi("Martin Morgan", "shout")
#' hi("Martin Morgan", "whisper")
```

Update and install the package, and test it

```{r, eval=FALSE}
devtools::document(pkg)
devtools::install(pkg)
example(hi)
```

Review and commit changes and push to github

```
## update Version in DESCRIPTION
git diff
git commit -a
git push
```

# 2. Classical, tidy, and rich approaches to data representation and analysis

- For session 2, please install [dplyr][], [data.table][],
  [reshape2][] [ggplot2][] and [SummarizedExperiment][]

        install.packages(
            c("dplyr", "data.table", "reshape2", "ggplot2"),
            repos="https://cran.r-project.org"
        )
        source("https://bioconductor.org/biocLite.R")
        biocLite("SummarizedExperiment")

- Please also download the following plain-text files:
  [ALL-expression.csv][], [ALL-phenoData.csv][]

'Classic' data is represented as a `data.frame()` with
samples as rows and features as columns.

```{r, eval=FALSE}
pdatafl = "~/a/Elbo/ALL-phenoData.csv"
exprfl = "~/a/Elbo/ALL-expression.csv"
classic <- input_classic(pdatafl, exprfl)
classic[1:5, c(1:3, 22:24)]
```

One of the things we did was to plot the distribution of 
expression values across genes

```{r classic-exprs-hist, eval=FALSE}
hist(colMeans(classic[, -(1:22)]))
```

# 3. Understanding classes and methods

# 4. Unit tests and other programming best practices

# 5. Writing efficient code

- For session 5, please install [microbenchmark][], [memoise][], and
  optionally [inline][]
  
        install.packages(
            c("microbenchmark", "memoise", "inline")
            repos="https://cran.r-project.org"
        )

# 6. Reports, shiny applications, and interactive communication

- For this session, please install [rmarkdown][] and [shiny][].

[R]: https://cran.r-project.org/
[Bioconductor]: https://bioconductor.org/
[RStudio]: https://www.rstudio.com/products/rstudio/download3/
[git]: https://git-scm.com/downloads
[devtools]: https://cran.r-project.org/package=devtools
[dplyr]: https://cran.r-project.org/package=dplyr
[data.table]: https://cran.r-project.org/package=data.table
[reshape2]: https://cran.r-project.org/package=reshape2
[ggplot2]: https://cran.r-project.org/package=ggplot2
[microbenchmark]: https://cran.r-project.org/package=microbenchmark
[memoise]: https://cran.r-project.org/package=memoise
[inline]: https://cran.r-project.org/package=inline
[rmarkdown]: https://cran.r-project.org/package=rmarkdown
[shiny]: https://cran.r-project.org/package=shiny
[SummarizedExperiment]: https://bioconductor.org/packages/SummarizedExperiment
[ALL-expression.csv]: https://github.com/Bioconductor/BiocIntroRPCI/raw/master/inst/extdata/ALL-expression.csv
[ALL-phenoData.csv]: https://github.com/Bioconductor/BiocIntroRPCI/raw/master/inst/extdata/ALL-phenoData.csv

[1]: https://cran.r-project.org/package=devtools
[roxygen2]: https://cran.r-project.org/package=roxygen2
[github]: https://github.com
