---
title: "Workshop: Building Packages"
author: "Martin Morgan"
date: "31 January, 2017"
vignette: >
  %\VignetteIndexEntry{Workshop: Intermediate R Software Development}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, echo=FALSE}
suppressPackageStartupMessages({
    library(devtools)
    library(roxygen2)
})
```

# 2. Classical, tidy, and rich approaches to data representation and analysis

- For session 2, please install [dplyr][], [data.table][],
  [reshape2][] [ggplot2][] and [SummarizedExperiment][]

        install.packages(
            c("dplyr", "reshape2", "ggplot2"),
            repos="https://cran.r-project.org"
        )
        source("https://bioconductor.org/biocLite.R")
        biocLite("SummarizedExperiment")

- Please also download the following plain-text files:
  [ALL-expression.csv][], [ALL-phenoData.csv][]

'Classic' data is represented as a `data.frame()` with
samples as rows and features as columns.

```{r, eval=FALSE}
pdatafl = "~/a/Elbo/ALL-phenoData.csv"
exprfl = "~/a/Elbo/ALL-expression.csv"
classic <- input_classic(pdatafl, exprfl)
classic[1:5, c(1:3, 22:24)]
```

One of the things we did was to plot the distribution of
expression values across genes

```{r classic-exprs-hist, eval=FALSE}
hist(colMeans(classic[, -(1:22)]))
```

# 5. Writing efficient code

- For session 5, please install [microbenchmark][], [memoise][], and
  optionally [inline][]

        install.packages(
            c("microbenchmark", "memoise", "inline")
            repos="https://cran.r-project.org"
        )

# 6. Reports, shiny applications, and interactive communication

- For this session, please install [rmarkdown][] and [shiny][].

[R]: https://cran.r-project.org/
[Bioconductor]: https://bioconductor.org/
[RStudio]: https://www.rstudio.com/products/rstudio/download3/
[git]: https://git-scm.com/downloads
[github]: https://github.com
[devtools]: https://cran.r-project.org/package=devtools
[roxygen2]: https://cran.r-project.org/package=roxygen2

[dplyr]: https://cran.r-project.org/package=dplyr
[SummarizedExperiment]: https://bioconductor.org/packages/SummarizedExperiment
[data.table]: https://cran.r-project.org/package=data.table
[reshape2]: https://cran.r-project.org/package=reshape2
[ggplot2]: https://cran.r-project.org/package=ggplot2

[ALL-expression.csv]: https://github.com/Bioconductor/BiocIntroRPCI/raw/master/inst/extdata/ALL-expression.csv
[ALL-phenoData.csv]: https://github.com/Bioconductor/BiocIntroRPCI/raw/master/inst/extdata/ALL-phenoData.csv

[microbenchmark]: https://cran.r-project.org/package=microbenchmark
[memoise]: https://cran.r-project.org/package=memoise
[inline]: https://cran.r-project.org/package=inline

[rmarkdown]: https://cran.r-project.org/package=rmarkdown
[shiny]: https://cran.r-project.org/package=shiny
